### 2023-06-02 submission/manuscript.tex
Editing

In 2014, McMurdie and Holmes published the provocatively titled, ``Waste
not, want not: why rarefying microbiome data is inadmissible''. The
claims of their study have significantly altered how microbiome
researchers control for the unavoidable uneven sequencing depths that
are inherent in modern 16S rRNA gene sequencing. Confusion over the
distinction between the definitions of rarefying and rarefaction
continues to cloud the interpretation of their results. More
importantly, the authors made a variety of choices in designing and
analyzing their simulations that were potentially problematic. I
identified 11 factors that could have compromised the results of the
original study. I reproduced the original simulation results and
assessed the impact of those 11 factors on the underlying conclusion
that rarefying data is inadmissible. Throughout, the design of the
original study made choices that caused rarefying and rarefaction to
appear to perform worse than they truly did. Most important were the
approaches used to assess ecological distances, the removal of samples
with low sequencing depth, and not accounting for conditions where
sampling effort is confounded with treatment group. Although the
original study criticized rarefying for the arbitrary removal of valid
data, repeatedly rarefying data many times (i.e., rarefaction)
incorporates all the data. In contrast, it is the removal of rare taxa
that would appear to remove valid data. Overall, I show that rarefaction
is the most robust approach to control for uneven sampling effort when
considered across a variety of alpha and beta diversity metrics.

@@

Microbiome studies that use amplicon sequencing to characterize multiple
samples use PCR to amplify 16S rRNA gene fragments using primers with
distinct barcodes or index sequences (1--3). These barcodes allow
researchers to pool PCR products and then deconvolute the resulting
sequence data based on the barcode sequences. Despite researchers' best
efforts to generate equimolar pools of PCR products, it is common to
observe more than 10-fold variation in the number of sequences per
sample (4). Researchers desire strategies to minimize uneven sequencing
depth and thus need methods to control for this unevenness in their
analyses. Of course, uneven sampling is not unique to microbiome
research and is a challenge faced by all community ecologists (5, 6).
Common approaches to control for the effects of uneven sequencing depths
include use of proportional abundance (i.e., relative abundance),
scaling of counts, parameter estimation, and rarefaction (7--23).

@@

This definition aligns with how rarefaction was originally defined for
comparing richness (i.e., the number of taxa in a community) across
communities when communities are sampled to different depths (5, 6).
With this more general approach to the definition of rarefaction,
rarefaction can be performed using any alpha or beta diversity metric.
This strategy has been widely used by my research group and others and
is available in the mothur software package using commands such as
\texttt{summary.single}, \texttt{rarefaction.single},
\texttt{phylo.diversity}, and \texttt{dist.shared}. The procedure
outlined above could also be used for hypothesis tests of differential
abundance; however, consideration is needed to synthesize the results of
these tests across a large number of replications.

@@

For the WNWN cluster analysis, sampling distributions for the two
treatment groups were generated using human fecal and ocean data
originally take from the GlobalPatterns dataset (L129) (2). To generate
the fecal and ocean distributions, the authors included any operational
taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3
ocean samples (L60 and L137). The OTUs were sorted by how many of the 7
samples the OTUs were observed in followed by their total abundance
across all 7 samples (L139). From this sorted list they obtained the
identifiers of the first 2,000 OTUs (L66). Returning to the 7 samples
they selected the data for the corresponding 2,000 OTUs and pooled the
OTU abundances of the fecal and ocean samples separately to create two
sampling distributions (L144, L159-160, L197-198). Next, the fecal and
ocean distributions were mixed in 8 different fractions to generate two
community types that differed by varying effect sizes (i.e., 1.00, 1.15,
1.25, 1.50, 1.75, 2.00, 2.50, and 3.50; L170-195, L220); an effect size
of 1.00 generated a null model with no difference between the treatment
groups. To simulate the variation in sequencing depth across the 80
samples, they normalized the number of sequences from each of the 26
samples in the GlobalPatterns dataset so that the median number of
sequences (Ñ\textsubscript{L}) across the samples had 1,000, 2,000,
5,000, or 10,000 sequences (L324-325). They then randomly sampled the 26
normalized sequencing depths, with replacement, to generate 80 sampling
depths. From each community type, they simulated 40 samples by sampling
to the desired number of sequences (L73, L230-233 and L326-327). Each
simulation condition was repeated 5 times (L85). Finally, they removed
rare and low prevalence OTUs in two steps. First, they removed any OTUs
whose total abundance was less than 3 across all 80 samples and that did
not appear in at least 3 samples (L368-386). Second, they removed any
OTUs that did not have more than 1 sequence in more than 5\% of the 80
samples (i.e., 4 samples) and that did not have a total abundance across
the 80 samples greater than one half of the number of samples in each
community type (i.e., 20) (L523-538, L551). Simulation A consisted of
160 simulations (8 effect sizes x 4 median sampling depths x 5
replicates = 160 simulations).

@@

In the 10 years since WNWN was published, sequencing technology has
advanced and sequence collections have grown considerably. For more
modern datasets, it would be reasonable to expect a median number of
sequences larger than 10,000 (see Table 1 of (4)). Therefore, I included
an additional median depth of sampling value of 50,000 sequences with
WNWN's four median sequencing sampling depths (i.e., 1,000, 2,000,
5,000, 10,000). Additional sequencing coverage would be expected to
result in more robust distance values since there would be more
information represented in the data. Indeed, the added sampling depth
showed higher accuracy values at lower effect sizes for the combinations
of normalization methods and distance calculations (Figure 3). Increased
sequencing coverage also resulted in improved clustering accuracy for
lower effect sizes when the library size minimum quantile was decreased
(Figure 4). I will revisit the choice of the library size minimum
quantile below.

@@

In WNWN, the authors quantified the tradeoff between median sequencing
depth, the number of samples removed below the threshold, and clustering
accuracy (WNWN's Figure 5, my Figure 4). Although the optimal threshold
varied by distance metric, normalization method, and sequencing depth,
they removed samples whose number of sequences was less than the 15th
percentile (L404-419). They acknowledged that this screening step, which
was only used with subsampling, would decrease clustering accuracy
putting it at a relative disadvantage to the other methods (page 5,
column 1, last paragraph). Therefore, it was not surprising that the
peak clustering accuracy for their subsampled data was at 85\%. Because
the true best result would not known \emph{a priori} in an actual
microbiome study, it would be impossible for researchers to conduct a
sensitivity analysis comparing the tradeoffs between sequencing depth,
sample number, and clustering accuracy to select a sampling depth for
their analysis without the risk of P-hacking. The differences in
clustering accuracy between subsampling and rarefaction with PAM and
K-means clustering indicated that it was necessary to reassess the
tradeoff between the library size minimum quantile and clustering
accuracy. When using rarefaction, K-means clustering, and only
considering conditions with 2,000 or more sequences, there was not a
condition where setting a higher threshold resulted in a better accuracy
than using all of the samples (Figure 8). These results showed that for
modern sequencing depths, using the full datasets with rarefaction and
K-means clustering resulted in accuracies that were better than those
observed when removing the smallest 15\% of the samples from each
simulated dataset. When the WNWN Figure 4 was recast with these
approaches, rarefaction performed at least as well as any of the other
transformations with each distance calculation (Figure 9).

@@

As described above, the sequencing depths used in the 26 GlobalPatterns
datasets were used as the distribution to create sequencing depths for
the 80 samples that were generated in each simulation. The
GlobalPatterns datasets had a mean of 1,085,256.8 sequences and a median
of 1,106,849 sequences per dataset (Figure 10). The datasets ranged in
sequencing depth between 58,688 and 2,357,181 sequences for a 40.16-fold
difference. Rather than representing a typically observed distribution
of sequencing depths that would be skewed right (see Figure S1 from
(4)), the sampling distribution was normally distributed (Shapiro-Wilk
test of normality, P=0.57). From these simulations it is unclear how
sensitive the various normalizations and distance calculations were to a
more realistic skewed distribution. A second limitation of this sampling
distribution is that it only contained 26 unique sampling depths such
that each sampling depth would have been re-used an average of 3.08
times in each simulation. Yet, it is unlikely for a real sequence
collection to have duplicate sequencing depths. To reassess the WNWN
results in the context of a more typical distribution of sample sizes, I
created a new set of simulations to test the effect of the shape of the
distribution on the results. I created a simple sequencing depth
distribution where there were 80 depths logarithmically distributed
between the minimum and maximum sequencing depths of the GlobalPatterns
dataset (Figure 10). The median of this distribution was 372,040 and the
mean was 629,824.8. When I regenerated the WNWN's Figures 4 and 5 using
the log-scaled sequencing effort distribution, the differences in
normalization methods were more apparent (Figure 11 and 12). For each of
the distance calculators, rarefaction to the size of the smallest
dataset yielded accuracies that were at least as good as the other
methods across effect sizes and median sequencing depths. The difference
was most pronounced at smaller effect sizes and sequencing depths. When
comparing the performance of rarefaction across distance calculators for
different effect sizes, sequencing depths and size of smallest sample,
the accuracies I observed using the log-scaled sample sizes was at least
as good as those obtained using the GlobalPatterns-based distribution
(Figure 9).

@@

I and others have observed that not using rarefaction can lead to
falsely detecting differences between communities when sampling effort
is confounded with the treatment group (25, 40). Previous analyses
showed that in these situations rarefaction did the best job of
controlling the rates of false detection (i.e., Type I errors) and
maintaining the statistical power to detect differences (i.e., 1-rate of
Type II errors) of differences between groups of samples. Such
situations have been observed when comparing communities at different
body parts where one site is more likely to generate contaminating
sequence reads from the host (41). To determine whether this result was
replicated with the WNWN simulation framework, I created a sampling
distribution where sequencing depth was fully confounded with the
treatment group using both the GlobalPatterns and Log-scaled sequence
distributions. To confound the sequencing depth, sequencing depths from
one treatment group were drawn from below the median number of sequences
of the sampling distribution and those for the second treatment group
were from above the median. To assess the risk of falsely detecting
clusters, I compared the clustering accuracies using an effect size of
1.00 using both the confounded and randomized sampling distributions
(see rows 1 and 2 of Figure 14). The samples should have only been
assigned to one cluster; however, each of the clustering methods forced
the samples into two clusters. So, when there are two groups of 40
samples that do not differ, the best a method could do would be to
correctly assign 41 of the 80 samples for an accuracy of 0.51. The false
detection risk did not vary by method when the sequencing depth was
randomized across treatment groups. Yet, when the sequencing depth was
confounded with treatment group, rarefaction was the most consistent
normalization method for controlling the generation of spurious
clusters. At larger effect sizes the ability to correctly identify two
clusters increased when the sequencing depth was confounded with
treatment group (Figure 14). At the effect size of 1.15, the rarefied
data generated the highest accuracy clusters regardless of whether the
sequencing depths were confounded. Although the level of confounding in
this simulation was extreme, it highlights the ability of rarefaction to
control the false detection rate and the ability to correctly detect
clusters relative to the other normalization methods.

@@

Rarefaction was originally proposed as a method for controlling uneven
sampling effort when comparing community richness values (5, 6). Thus it
was surprising that WNWN did not consider the effect of the proposed
normalizations on alpha-diversity metrics such as richness or Shannon
diversity. Therefore, for each of the normalizations, I compared the
richness and diversity of the two treatment groups. The DESeq Variance
Stabilization normalized data were not included because the
normalization produced negative values which were not compatible with
calculations of richness or Shannon diversity. Also, data from the
Upper-Quartile Log-Fold Change normalization were not used for richness
calculations since the normalization returned the same richness values
for each sample regardless of the treatment group. I assessed whether
the alpha diversity was significantly differente between the treatment
groups for each iteration using the non-parametric Wilcoxon two-sampled
test. I compared the risk of committing Type I errors and the power to
detect differences by the different normalizations (Figures 16 and 17).
For these analyses, I used the GlobalPatterns data with the random and
confounded distribution of sequencing depths. Similar to the results in
points 9 and 10, with the exception of rarefaction, the simulations
using a confounded sequence depth distribution resulted in all of the
replicates having a significant test. The power to detect differences in
richness and diversity at effect sizes of 1.15 and greater with
rarefaction was at least as high as any of the other normalizations.

@@

The conclusions from WNWN have had a lasting impact on how researchers
analyze microbiome sequence data. As I have demonstrated using WNWN's
simulation framework, their claims are not supported. The most important
points that lead to the difference in our conclusions include the choice
of clustering algorithm and arbitrarily selecting a sequencing effort
threshold that was used to remove samples. Furthermore, the decision to
evaluate the effectiveness of normalization methods based on clustering
samples adds a layer of analysis that has been controversial and not
widely used. Ultimately, the authors choice of the word ``rarefying''
has sewn confusion in the field because it is often used in place of the
word ``rarefaction''. As I have demonstrated there were numerous choices
throughout the original study that resulted in rarefying/subsampling
looking worse than it truly is. In fact, when the data of the current
study are taken as a whole, rarefaction is the preferred approach. Short
of obtaining the exact same number of sequences from each sample,
rarefaction remains the best approach to control for uneven sampling
effort when analyzing alpha and beta diversity metrics. Although beyond
the scope of the current study, the same principles would support using
rarefaction to control for uneven sampling effort in shotgun metagenomic
sequencing analyses.

Beyond the discussion of whether rarefaction is appropriate for
analyzing microbiome data it is worth commenting on WNWN's advice to use
DESeq's Variance Stabilization or edgeR's Upper-Quartile Log-fold Change
normalization strategies. These methods have been adopted from gene
expression analysis to microbiome analysis. Gene expression analysis
implicitly assumes that all samples have the same genes since. While
this might work in comparing healthy and diseased tissues from a cohort
of patients, it does not generalize to those patients' microbiota.
Microbial populations are highly patchy in their distribution. Thus, a
zero count for gene expression is more likely to represent a gene below
the limit of detection whereas a zero count for a microbiome analysis is
more likely to represent the true absence of the OTU. In fact, zeroes
are less common in gene expression analyses. But because microbiome
studies have so many zeroes, it is necessary to add a pseudo-count to
all OTUs for both the edgeR and DESeq-based normalizations (L443 and
L487, respectively). In WNWN, a pseudo-count of 1 was used. However,
this value is arbitrary and the results can vary based on the choice of
pseudo-count and the patchiness of the communities being analyzed. Since
WNWN was published, compositional approaches have been proposed to
account for uneven sampling and to provide improved interpretability
(14, 16--18, 46, 47). However, these methods also often require the use
of pseudo-counts and are not actually insensitive to uneven sampling
(40). Rarefaction is preferred to these alternatives.

@@

Although the authors claim that ``Rarefying counts requires an arbitrary
selection of a library size minimum that affects downstream inference''
(page 8, column 1, point 3), in actual microbiome studies the selection
of a sampling depth is not as arbitrary as the authors claim. Rather, to
avoid ``P-hacking'', researchers pick a set of criteria where they will
include or exclude samples prior to testing their data. Examples of
criteria might include the presence of a large gap in the sequencing
effort distribution, a desire to include poorly sequenced controls, or
the \emph{a priori} stipulation of a minimum sequencing effort. To
mitigate concerns of arbitrary or engineered minimum library sizes,
researchers should indicate the rationale for the threshold they
selected.

@@

In a parallel set of analyses I have used alternative simulation and
evaluation strategies to look more closely at rarefaction and its
alternatives (40) and the practice of filtering low abundance sequences
(4). The results of those studies are similar to this analysis. These
studies demonstrate that it is actually sequence and OTU filtering that
``requires the omission of available valid data'' and that rarefaction
is the only available method for controlling the effects of uneven
sampling. Far from being inadmissible, rarefaction of unfiltered
datasets yields the most robust results.

@@

Next, each simulation was further processed by filtering rare OTUs, to
normalize uneven sampling depths, and to calculate ecological distances
between samples. Again, the code contained within
\texttt{simulation-cluster-accuracy-server.Rmd} was spilt into
individual R scripts and modified to make the generation of the
simulated communities more modular and scalable within the
\texttt{nr-s1} Conda environment. First, each simulation was either
filtered to remove rare OTUs or left unfiltered. As was done in the WNWN
study, filtering was performed in two steps: (1) any OTUs whose total
abundance was less than 3 across all 80 samples and that did not appear
in at least 3 samples and (2) any OTU that did not have more than 1
sequence in more than 5\% of the 80 samples (i.e., 4 samples) and that
did not have a total abundance across the 80 samples greater than one
half of the number of samples in each community type (i.e., 20) were
removed. Second, OTU data was normalized by one of six approaches: (1)
non-normalized raw counts, (2) relative abundance, (3) DESeq Variance
Stabilization, (4) Upper-Quartile Log-Fold Change normalization using
edgeR, (5) a single subsampling to a common number of sequences, (6)
rarefaction with 100 randomizations to a common number of sequences.
Although the data were not presented, the WNWN code and my code also
normalized the counts using trimmed mean of M-values (TMM) and relative
log expression (RLE) in edgeR. The level of subsampling and rarefaction
was selected based on the size of the sample whose sequencing depth was
at the 0 (i.e., included all of the samples), 5, 10, 15, 20, 25, and
40th percentiles. Finally, the normalized simulated datasets were used
to calculate pairwise distances between the 80 samples using seven
different calculations: Bray-Curtis, Euclidean, Poisson, Mean Squared
Difference, Unweighted UniFrac, Weighted Unifrac, and Biological
Coeffcient of Variation (BCV); however, data using BCV were not
discussed in this or the WNWN analysis. For the rarefaction normalized
data, each of the 100 subsampled datasets were used to calculate a
distance matrix. The mean of the 100 distance matrices was used as the
distance matrix for the rarefaction data. As indicated in the Results
section, there were combinations of normalization and distance methods
that were not compatible (e.g., DESeq Variance Stabilization
normalization with Bray-Curtis distances). In such cases, distance
matrices were coded as \texttt{NA}. My choice of combinations of
normalizations and distances to visualize was determined by the WNWN
analysis except where noted. For each simulated dataset there were 280
possible combinations of filtering, normalization, and distance methods
(2 filtering methods x 20 normalization methods x 7 distance methods).

>>>>

In 2014, McMurdie and Holmes published the provocatively titled, ``Waste
not, want not: why rarefying microbiome data is inadmissible''. The
claims of their study have significantly altered how microbiome
researchers control for the unavoidable uneven sequencing depths that
are inherent in modern 16S rRNA gene sequencing. Confusion over the
distinction between the definitions of rarefying and rarefaction
continues to cloud the interpretation of their results. More
importantly, the authors made a variety of choices in designing and
analyzing their simulations that were potentially problematic. I
identified 11 factors that could have compromised the results of the
original study. I reproduced the original simulation results and
assessed the impact of those 11 factors on the underlying conclusion
that rarefying data is inadmissible. Throughout, the design of the
original study made choices that caused rarefying and rarefaction to
appear to perform worse than they truly did. Most important were the
approaches used to assess ecological distances, the removal of samples
with low sequencing depth, and not accounting for conditions where
sequencing effort is confounded with treatment group. Although the
original study criticized rarefying for the arbitrary removal of valid
data, repeatedly rarefying data many times (i.e., rarefaction)
incorporates all the data. In contrast, it is the removal of rare taxa
that would appear to remove valid data. Overall, I show that rarefaction
is the most robust approach to control for uneven sequencing effort when
considered across a variety of alpha and beta diversity metrics.

@@

Microbiome studies that use amplicon sequencing to characterize multiple
samples use PCR to amplify 16S rRNA gene fragments using primers with
distinct barcodes or index sequences (1--3). These barcodes allow
researchers to pool PCR products and then deconvolute the resulting
sequence data based on the barcode sequences. Despite researchers' best
efforts to generate equimolar pools of PCR products, it is common to
observe more than 10-fold variation in the number of sequences per
sample (4). Researchers desire strategies to minimize uneven sequencing
depth and thus need methods to control for this unevenness in their
analyses. Of course, uneven sequencing is not unique to microbiome
research and is a challenge faced by all community ecologists (5, 6).
Common approaches to control for the effects of uneven sequencing depths
include use of proportional abundance (i.e., relative abundance),
scaling of counts, parameter estimation, and rarefaction (7--23).

@@

This definition aligns with how rarefaction was originally defined for
comparing richness (i.e., the number of taxa in a community) across
communities when communities are sequenced to different depths (5, 6).
With this more general approach to the definition of rarefaction,
rarefaction can be performed using any alpha or beta diversity metric.
This strategy has been widely used by my research group and others and
is available in the mothur software package using commands such as
\texttt{summary.single}, \texttt{rarefaction.single},
\texttt{phylo.diversity}, and \texttt{dist.shared}. The procedure
outlined above could also be used for hypothesis tests of differential
abundance; however, consideration is needed to synthesize the results of
these tests across a large number of replications.

@@

For the WNWN cluster analysis, OTU abundance distributions for the two
treatment groups were generated using human fecal and ocean data
originally take from the GlobalPatterns dataset (L129) (2). To generate
the fecal and ocean distributions, the authors included any operational
taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3
ocean samples (L60 and L137). The OTUs were sorted by how many of the 7
samples the OTUs were observed in followed by their total abundance
across all 7 samples (L139). From this sorted list they obtained the
identifiers of the first 2,000 OTUs (L66). Returning to the 7 samples
they selected the data for the corresponding 2,000 OTUs and pooled the
OTU abundances of the fecal and ocean samples separately to create two
OTU abundance distributions (L144, L159-160, L197-198). Next, the fecal
and ocean distributions were mixed in 8 different fractions to generate
two community types that differed by varying effect sizes (i.e., 1.00,
1.15, 1.25, 1.50, 1.75, 2.00, 2.50, and 3.50; L170-195, L220); an effect
size of 1.00 generated a null model with no difference between the
treatment groups. To simulate the variation in sequencing depth across
the 80 samples, they normalized the number of sequences from each of the
26 samples in the GlobalPatterns dataset so that the median number of
sequences (Ñ\textsubscript{L}) across the samples had 1,000, 2,000,
5,000, or 10,000 sequences (L324-325). They then randomly sampled the 26
normalized sequencing depths, with replacement, to generate 80
sequencing depths. From each community type, they simulated 40 samples
by sampling to the desired number of sequences (L73, L230-233 and
L326-327). Each simulation condition was repeated 5 times (L85).
Finally, they removed rare and low prevalence OTUs in two steps. First,
they removed any OTUs whose total abundance was less than 3 across all
80 samples and that did not appear in at least 3 samples (L368-386).
Second, they removed any OTUs that did not have more than 1 sequence in
more than 5\% of the 80 samples (i.e., 4 samples) and that did not have
a total abundance across the 80 samples greater than one half of the
number of samples in each community type (i.e., 20) (L523-538, L551).
Simulation A consisted of 160 simulations (8 effect sizes x 4 median
sequencing depths x 5 replicates = 160 simulations).

@@

In the 10 years since WNWN was published, sequencing technology has
advanced and sequence collections have grown considerably. For more
modern datasets, it would be reasonable to expect a median number of
sequences larger than 10,000 (see Table 1 of (4)). Therefore, I included
an additional median depth of sequencing of 50,000 sequences with WNWN's
four median sequencing depths (i.e., 1,000, 2,000, 5,000, 10,000).
Additional sequencing coverage would be expected to result in more
robust distance values since there would be more information represented
in the data. Indeed, the added sequencing depth showed higher accuracy
values at lower effect sizes for the combinations of normalization
methods and distance calculations (Figure 3). Increased sequencing
coverage also resulted in improved clustering accuracy for lower effect
sizes when the library size minimum quantile was decreased (Figure 4). I
will revisit the choice of the library size minimum quantile below.

@@

In WNWN, the authors quantified the tradeoff between median sequencing
depth, the number of samples removed below the threshold, and clustering
accuracy (WNWN's Figure 5, my Figure 4). Although the optimal threshold
varied by distance metric, normalization method, and sequencing depth,
they removed samples whose number of sequences was less than the 15th
percentile (L404-419). They acknowledged that this screening step, which
was only used with subsampling, would decrease clustering accuracy
putting it at a relative disadvantage to the other methods (page 5,
column 1, last paragraph). Therefore, it was not surprising that the
peak clustering accuracy for their subsampled data was at 85\%. Because
the true best result would not known \emph{a priori} in an actual
microbiome study, it would be impossible for researchers to conduct a
sensitivity analysis comparing the tradeoffs between sequencing depth,
sample number, and clustering accuracy to select a sequencing depth for
their analysis without the risk of P-hacking. The differences in
clustering accuracy between subsampling and rarefaction with PAM and
K-means clustering indicated that it was necessary to reassess the
tradeoff between the library size minimum quantile and clustering
accuracy. When using rarefaction, K-means clustering, and only
considering conditions with 2,000 or more sequences, there was not a
condition where setting a higher threshold resulted in a better accuracy
than using all of the samples (Figure 8). These results showed that for
modern sequencing depths, using the full datasets with rarefaction and
K-means clustering resulted in accuracies that were better than those
observed when removing the smallest 15\% of the samples from each
simulated dataset. When the WNWN Figure 4 was recast with these
approaches, rarefaction performed at least as well as any of the other
transformations with each distance calculation (Figure 9).

@@

As described above, the sequencing depths used in the 26 GlobalPatterns
datasets were used as the distribution to create sequencing depths for
the 80 samples that were generated in each simulation. The
GlobalPatterns datasets had a mean of 1,085,256.8 sequences and a median
of 1,106,849 sequences per dataset (Figure 10). The datasets ranged in
sequencing depth between 58,688 and 2,357,181 sequences for a 40.16-fold
difference. Rather than representing a typically observed distribution
of sequencing depths that would be skewed right (see Figure S1 from
(4)), the sequencing depth distribution was normally distributed
(Shapiro-Wilk test of normality, P=0.57). From these simulations it is
unclear how sensitive the various normalizations and distance
calculations were to a more realistic skewed distribution. A second
limitation of this sequencing depth distribution is that it only
contained 26 unique sequencing depths such that each sequencing depth
would have been re-used an average of 3.08 times in each simulation.
Yet, it is unlikely for a real sequence collection to have duplicate
sequencing depths. To reassess the WNWN results in the context of a more
typical distribution of sequencing depths, I created a new set of
simulations to test the effect of the shape of the distribution on the
results. I created a simple sequencing depth distribution where there
were 80 depths logarithmically distributed between the minimum and
maximum sequencing depths of the GlobalPatterns dataset (Figure 10). The
median of this distribution was 372,040 and the mean was 629,824.8. When
I regenerated the WNWN's Figures 4 and 5 using the log-scaled sequencing
effort distribution, the differences in normalization methods were more
apparent (Figure 11 and 12). For each of the distance calculators,
rarefaction to the size of the smallest dataset yielded accuracies that
were at least as good as the other methods across effect sizes and
median sequencing depths. The difference was most pronounced at smaller
effect sizes and sequencing depths. When comparing the performance of
rarefaction across distance calculators for different effect sizes,
sequencing depths and size of smallest sample, the accuracies I observed
using the log-scaled sample sizes was at least as good as those obtained
using the GlobalPatterns-based distribution (Figure 9).

@@

I and others have observed that not using rarefaction can lead to
falsely detecting differences between communities when sequencing effort
is confounded with the treatment group (25, 40). Previous analyses
showed that in these situations rarefaction did the best job of
controlling the rates of false detection (i.e., Type I errors) and
maintaining the statistical power to detect differences (i.e., 1-rate of
Type II errors) of differences between groups of samples. Such
situations have been observed when comparing communities at different
body parts where one site is more likely to generate contaminating
sequence reads from the host (41). To determine whether this result was
replicated with the WNWN simulation framework, I created a sequencing
depth distribution where sequencing depth was fully confounded with the
treatment group using both the GlobalPatterns and Log-scaled sequence
distributions. To confound the sequencing depth, sequencing depths from
one treatment group were drawn from below the median number of sequences
of the sequencing distribution and those for the second treatment group
were from above the median. To assess the risk of falsely detecting
clusters, I compared the clustering accuracies using an effect size of
1.00 using both the confounded and randomized sequencing distributions
(see rows 1 and 2 of Figure 14). The samples should have only been
assigned to one cluster; however, each of the clustering methods forced
the samples into two clusters. So, when there are two groups of 40
samples that do not differ, the best a method could do would be to
correctly assign 41 of the 80 samples for an accuracy of 0.51. The false
detection risk did not vary by method when the sequencing depth was
randomized across treatment groups. Yet, when the sequencing depth was
confounded with treatment group, rarefaction was the most consistent
normalization method for controlling the generation of spurious
clusters. At larger effect sizes the ability to correctly identify two
clusters increased when the sequencing depth was confounded with
treatment group (Figure 14). At the effect size of 1.15, the rarefied
data generated the highest accuracy clusters regardless of whether the
sequencing depths were confounded. Although the level of confounding in
this simulation was extreme, it highlights the ability of rarefaction to
control the false detection rate and the ability to correctly detect
clusters relative to the other normalization methods.

@@

Rarefaction was originally proposed as a method for controlling uneven
sequencing effort when comparing community richness values (5, 6). Thus
it was surprising that WNWN did not consider the effect of the proposed
normalizations on alpha-diversity metrics such as richness or Shannon
diversity. Therefore, for each of the normalizations, I compared the
richness and diversity of the two treatment groups. The DESeq Variance
Stabilization normalized data were not included because the
normalization produced negative values which were not compatible with
calculations of richness or Shannon diversity. Also, data from the
Upper-Quartile Log-Fold Change normalization were not used for richness
calculations since the normalization returned the same richness values
for each sample regardless of the treatment group. I assessed whether
the alpha diversity was significantly different between the treatment
groups for each iteration using the non-parametric Wilcoxon two-sampled
test. I compared the risk of committing Type I errors and the power to
detect differences by the different normalizations (Figures 16 and 17).
For these analyses, I used the GlobalPatterns data with the random and
confounded distribution of sequencing depths. Similar to the results in
points 9 and 10, with the exception of rarefaction, the simulations
using a confounded sequence depth distribution resulted in all of the
replicates having a significant test. The power to detect differences in
richness and diversity at effect sizes of 1.15 and greater with
rarefaction was at least as high as any of the other normalizations.

@@

The conclusions from WNWN have had a lasting impact on how researchers
analyze microbiome sequence data. As I have demonstrated using WNWN's
simulation framework, their claims are not supported. The most important
points that lead to the difference in our conclusions include the choice
of clustering algorithm and arbitrarily selecting a sequencing effort
threshold that was used to remove samples. Furthermore, the decision to
evaluate the effectiveness of normalization methods based on clustering
samples adds a layer of analysis that has been controversial and not
widely used. Ultimately, the authors choice of the word ``rarefying''
has sewn confusion in the field because it is often used in place of the
word ``rarefaction''. As I have demonstrated there were numerous choices
throughout the original study that resulted in rarefying/subsampling
looking worse than it truly is. In fact, when the data of the current
study are taken as a whole, rarefaction is the preferred approach. Short
of obtaining the exact same number of sequences from each sample,
rarefaction remains the best approach to control for uneven sequencing
effort when analyzing alpha and beta diversity metrics. Although beyond
the scope of the current study, the same principles would support using
rarefaction to control for uneven sequencing effort in shotgun
metagenomic sequencing analyses.

Beyond the discussion of whether rarefaction is appropriate for
analyzing microbiome data it is worth commenting on WNWN's advice to use
DESeq's Variance Stabilization or edgeR's Upper-Quartile Log-fold Change
normalization strategies. These methods have been adopted from gene
expression analysis to microbiome analysis. Gene expression analysis
implicitly assumes that all samples have the same genes since. While
this might work in comparing healthy and diseased tissues from a cohort
of patients, it does not generalize to those patients' microbiota.
Microbial populations are highly patchy in their distribution. Thus, a
zero count for gene expression is more likely to represent a gene below
the limit of detection whereas a zero count for a microbiome analysis is
more likely to represent the true absence of the OTU. In fact, zeroes
are less common in gene expression analyses. But because microbiome
studies have so many zeroes, it is necessary to add a pseudo-count to
all OTUs for both the edgeR and DESeq-based normalizations (L443 and
L487, respectively). In WNWN, a pseudo-count of 1 was used. However,
this value is arbitrary and the results can vary based on the choice of
pseudo-count and the patchiness of the communities being analyzed. Since
WNWN was published, compositional approaches have been proposed to
account for uneven sequencing and to provide improved interpretability
(14, 16--18, 46, 47). However, these methods also often require the use
of pseudo-counts and are not actually insensitive to uneven sequencing
(40). Rarefaction is preferred to these alternatives.

@@

Although the authors claim that ``Rarefying counts requires an arbitrary
selection of a library size minimum that affects downstream inference''
(page 8, column 1, point 3), in actual microbiome studies the selection
of a sequencing depth is not as arbitrary as the authors claim. Rather,
to avoid ``P-hacking'', researchers pick a set of criteria where they
will include or exclude samples prior to testing their data. Examples of
criteria might include the presence of a large gap in the sequencing
effort distribution, a desire to include poorly sequenced controls, or
the \emph{a priori} stipulation of a minimum sequencing effort. To
mitigate concerns of arbitrary or engineered minimum library sizes,
researchers should indicate the rationale for the threshold they
selected.

@@

In a parallel set of analyses I have used alternative simulation and
evaluation strategies to look more closely at rarefaction and its
alternatives (40) and the practice of filtering low abundance sequences
(4). The results of those studies are similar to this analysis. These
studies demonstrate that it is actually sequence and OTU filtering that
``requires the omission of available valid data'' and that rarefaction
is the only available method for controlling the effects of uneven
sequencing. Far from being inadmissible, rarefaction of unfiltered
datasets yields the most robust results.

@@

Next, each simulation was further processed by filtering rare OTUs, to
normalize uneven sequencing depths, and to calculate ecological
distances between samples. Again, the code contained within
\texttt{simulation-cluster-accuracy-server.Rmd} was spilt into
individual R scripts and modified to make the generation of the
simulated communities more modular and scalable within the
\texttt{nr-s1} Conda environment. First, each simulation was either
filtered to remove rare OTUs or left unfiltered. As was done in the WNWN
study, filtering was performed in two steps: (1) any OTUs whose total
abundance was less than 3 across all 80 samples and that did not appear
in at least 3 samples and (2) any OTU that did not have more than 1
sequence in more than 5\% of the 80 samples (i.e., 4 samples) and that
did not have a total abundance across the 80 samples greater than one
half of the number of samples in each community type (i.e., 20) were
removed. Second, OTU data was normalized by one of six approaches: (1)
non-normalized raw counts, (2) relative abundance, (3) DESeq Variance
Stabilization, (4) Upper-Quartile Log-Fold Change normalization using
edgeR, (5) a single subsampling to a common number of sequences, (6)
rarefaction with 100 randomizations to a common number of sequences.
Although the data were not presented, the WNWN code and my code also
normalized the counts using trimmed mean of M-values (TMM) and relative
log expression (RLE) in edgeR. The level of subsampling and rarefaction
was selected based on the size of the sample whose sequencing depth was
at the 0 (i.e., included all of the samples), 5, 10, 15, 20, 25, and
40th percentiles. Finally, the normalized simulated datasets were used
to calculate pairwise distances between the 80 samples using seven
different calculations: Bray-Curtis, Euclidean, Poisson, Mean Squared
Difference, Unweighted UniFrac, Weighted Unifrac, and Biological
Coeffcient of Variation (BCV); however, data using BCV were not
discussed in this or the WNWN analysis. For the rarefaction normalized
data, each of the 100 subsampled datasets were used to calculate a
distance matrix. The mean of the 100 distance matrices was used as the
distance matrix for the rarefaction data. As indicated in the Results
section, there were combinations of normalization and distance methods
that were not compatible (e.g., DESeq Variance Stabilization
normalization with Bray-Curtis distances). In such cases, distance
matrices were coded as \texttt{NA}. My choice of combinations of
normalizations and distances to visualize was determined by the WNWN
analysis except where noted. For each simulated dataset there were 280
possible combinations of filtering, normalization, and distance methods
(2 filtering methods x 20 normalization methods x 7 distance methods).
